---
title: "Olotillo_map"
output: html_document
---

# Title: Plot map.
# Star date: October 2023 by Alicia Mastretta-Yanes.
# Previous modification: Mon 13/Oct/2025. Versoix, Switzerland (14:03) by Duhyadi Oliva-García.
# Last modification: Wed 22/Oct/2025. Versoix, Switzerland (13:57) by Duhyadi Oliva-García.
# Authors: Duhyadi Oliva-García & Alicia Mastretta-Yanes.

------------------------------------------------------------------------

# INITIAL PROJECT SETUP  
```{r setup, include=FALSE}
# --- Load the package here (for reproducible routes) ----------------------
if (requireNamespace("here", quietly = TRUE)) {
  setwd(here::here())
  message("📂  Working directory set to: ", getwd())
} else {
  stop("⚠️ The 'here' package is not installed. Execute install.packages('here')")
}
```

------------------------------------------------------------------------

# LOAD LIBRARIES
```{r}
library(dplyr)
library(ggplot2)
library(sf)
```

# LOAD DATA: read shape and points
Occurrence points for Olotillo, were collected in the field for, 2018-2019: Duhyadi Oliva-García (Chiapas), Alicia Mastretta-Yanes(Chiapas), Bulmaro Coutiño-Estrada (Chiapas), Flavio Aragón-Cuevas (Oaxaca), Noé Orlando-Gómez (Guerrero), Víctor Antonio Vidal Martínez (Nayarit) and Hugo Perales-Rivera (Chiapas-different year). Also include samples from CIMMYTM.
```{r}
# --- Load spatial layers and data -----------------------------------------
# here::here. It's the safe way to call package here.
mexico <- st_read(here::here("shape","destdv1gw" ,"destdv1gw.shp")) # Mexico shape
distribution <- st_read(here::here("shape","oloti_7cgw" ,"oloti_7cgw.shp")) # Olotillo distribution shape
olotillo.PGMN <- read.csv(here::here("meta","olotillo_PGMN_siagro.csv")) # Olotillo, Proyecto Global Maíces Nativos (PGMN), CONABIO.Global Maize Project (CONABIO, 2011).
olotillo.points <- read.csv(here::here("meta", "olotillo_points_meta_clean.csv")) # Olotillo, Doctoral Project Evolutionary Populations
```

# PLOT
## MAP #1 
## ------------------------------------------------------------------
## 🗺️ Base map of Mexico with distribution...
## ------------------------------------------------------------------
Distribution of Olotillo according to Perales & Golicher (2014) + points for Olotillo the Global Maize Project (CONABIO, 2011). + accessions sampled for this study by race.
```{r}
# --- Create the map -----------------------------------------
map1 <- ggplot() +
  geom_sf(data = mexico, colour = "grey", fill = NA) + # shape mexico
  geom_sf(data = distribution, colour = "blue", fill = NA, size = 0.2) + # shape distribution
  geom_point(data = olotillo.PGMN, aes(x = longitud, y = latitud), size = 0.5, colour = "black") +
  geom_point(data = olotillo.points, aes(x = longitude, y = latitude, shape = Race, color = Race), size = 1.5) +
  scale_shape_manual(values = c(15, 17, 16, 18)) +
  scale_color_manual(values = c("purple", "green", "red", "orange")) +
  theme_void()
# --- Show map -----------------------------------------
map1
```

## MAP #2
## ------------------------------------------------------------------
## 🗺️ Base map of Mexico with distribution and points of PGMN (Olotillo)
## ------------------------------------------------------------------
Climatic interpolation of Olotillo distribution according to Perales & Golicher (2014) + sampling points for Olotillo from the Global Maize Project (CONABIO, 2011). 
**Figure 1.Distribution and environmental conditions where Olotillo is grown.** 
**a)** Purple color shows the climatic interpolation of Olotillo distribution, restricted to the main altitude where it is grown, according to Perales & Golicher (2014). Black points are the sampling points of Olotillo from the Global Maize Project (GMP: CONABIO, 2011).
```{r}
# --- Create the map -----------------------------------------
map2 <- ggplot() +
  geom_sf(data = mexico, colour = "grey", fill = NA, linewidth = 0.3) + # base polygon of Mexico
  geom_sf(data = distribution, colour = "purple", fill = NA, linewidth = 0.6) + # distribution polygon Olotillo 
  geom_point(                          # points of PGMN
    data = olotillo.PGMN,
    aes(x = longitud, y = latitud),
    colour = "black",
    size = 0.8
  ) + 
  theme_void()                         
# --- Show map -----------------------------------------
map2
```

## MAP #3
## ------------------------------------------------------------------
## 🗺️ Map of Mexico highligthing states that were sampled (same colors than PCA)
## ------------------------------------------------------------------
```{r}
# --- Color by states -----------------------------------------
mexico_sf <- st_as_sf(mexico, region = "ENTIDAD")
lapply(mexico_sf, class)
print(mexico_sf)
```
### Convert shape to data frame: extract coordinates and attributes in data frame format
```{r}
# ---- Extract coordinates and correctly join them with the attributes ---------------------
coords <- st_coordinates(mexico) %>% as.data.frame()
coords$ID <- as.numeric(rownames(coords))   # create temporary ID
# ---- Convert geometries to data frame (maintaining attributes) ---------------------------
mexico_df <- mexico %>%
  st_drop_geometry() %>%                    # delete geometry (maintains attributes)
  mutate(ID = row_number()) %>%             # create reference ID
  right_join(coords, by = "ID")             # join with coordinates

                                                                
head(mexico_df)                             # check result 
```
### Pop data
```{r}
# --- Read the file plate1_fullmeta.txt -----------------------------------------
pop.data <- read.table(
  here::here("meta", "plate1_fullmeta.txt"),
  sep = "\t",
  header = TRUE,
  stringsAsFactors = FALSE
)
pop.data <- pop.data[order(pop.data$sample_name),]  # order alphabetically pop.data
pop.data <- pop.data[-1,]                           # remove the first element, (DNA BLANK)
dim(pop.data)                                       # check n samples
```
### Make a vector 
```{r}
# --- Make vector of states excluding where we sampled olotillos --------------------------
sampled.states<-unique(pop.data$state)
sampled.states
# To get a names in the data frame
unique(mexico_sf$ENTIDAD) # to get a names in the data frame
# Change to they way 
sampled.states<-c("CAMPECHE", "CHIAPAS", "GUERRERO", "HIDALGO",
                  "NAYARIT", "OAXACA", "QUINTANA ROO", "SAN LUIS POTOSI",
                  "VERACRUZ DE IGNACIO DE LA LLAVE")
not.sampled<-unique(mexico_sf$ENTIDAD)[!unique(mexico_sf$ENTIDAD) %in% sampled.states] 
```

```{r}
# --- Create the map showing only sampled states -----------------------------------------
map3 <- ggplot() +
  geom_sf(
    data = mexico_sf,
    aes(fill = ENTIDAD),
    colour = "grey40",
    linewidth = 0.3
  ) +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_manual(
    breaks = c(sampled.states, not.sampled),
    values = c(
      "grey60",     # base
      "#75a842",    # verde
      "#36dee6",    # turquesa
      "#52bd85",    # verde medio
      "#b94656",    # rojo
      "#6780d8",    # azul
      "#c69c39",    # dorado
      "#b6b638",    # oliva
      "#89863c",    # marrón verdoso
      rep("white", 23)
    )
  ) 

map3
```

# Add sampling points of Olotillo from PGMN
```{r}
# --- Create the map showing only sampled states -----------------------------------------
map3b <- ggplot() +
  geom_sf(
    data = mexico_sf,
    aes(fill = ENTIDAD),
    colour = "grey40",
    linewidth = 0.3
  ) +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_manual(
    breaks = c(sampled.states, not.sampled),
    values = c(
      "grey60",     # base
      "#75a842",    # verde
      "#36dee6",    # turquesa
      "#52bd85",    # verde medio
      "#b94656",    # rojo
      "#6780d8",    # azul
      "#c69c39",    # dorado
      "#b6b638",    # oliva
      "#89863c",    # marrón verdoso
      rep("white", 23)
    )
  ) +
  geom_point(
    data = olotillo.PGMN,
    aes(x = longitud, y = latitud),
    colour = "black",
    size = 0.6
  ) +
  geom_point(
    data = olotillo.points,
    aes(x = longitude, y = latitude, shape = Race, color = Race),
    size = 1.6
  ) +
  scale_shape_manual(values = c(15, 17, 16, 18)) +
  scale_color_manual(values = c("purple", "green", "blue", "orange")) +
  theme_void() +
  theme(legend.position = "bottom") +
  guides(
    shape = guide_legend(title = "Race"),
    color = guide_legend(title = "Race")
  )
map3b
```







