---
title: "Untitled"
output: html_document
date: "2025-08-12"
---

# Title: Supplementary Table 3. Genetic diversity values by accession.
# Start date: August 12-2025. Wakerhust.
# Previous modification: Mon 27/Oct/2025. Verxoix, Switzerland (15:00 PM).   
# Last modification:     Tue 28/Oct/2025. Verxoix, Switzerland (13:00 PM). 
# Authors: Duhyadi Oliva-García & Alicia Mastretta-Yanes.
--------

# WORKING DIRECTORY 
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/home/duhyadi/Documents/paper1_final_tipodryad")
```

# LIBRARIES
```{r}
library(doBy)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(readr)
library(stringr)
```

# LOAD FILE: generated by vcftools 
```{r}
df <- read.csv("tables_supplementary_diversity/meta/data_diversity_ago2025")
```
# Select the Olotillo race and variables of interest
```{r}
df_olotillo <- df[df$Race == "Olotillo",]
df_sel <- df_olotillo %>% 
  select(accession_ID, F, HETob, HETex, HOMex)
```
# Group by accession_ID
```{r}
summary <- df_sel %>% 
  group_by(accession_ID) %>% 
  summarise(
    F_mean        = round(mean (F, na.rm = TRUE),3),
    F_sd          = round(sd   (F, na.rm = TRUE),3),
    HETob_mean    = round(mean (HETob , na.rm = TRUE),3),
    HETob_sd      = round(sd   (HETob , na.rm = TRUE),3),
    HETex_mean    = round(mean (HETex , na.rm = TRUE),3),
    HETex_sd      = round(sd (HETex , na.rm = TRUE),3),
    HOMex_mean    = round(mean (HOMex , na.rm = TRUE),3),
    HOMex_sd      = round(sd (HOMex , na.rm = TRUE),3),
    n = n(),
    .groups = "drop"
  ) %>% 
  arrange(accession_ID)
```
# Print summary
```{r}
print(summary, n = 50)
```
# Tail summary
```{r}
tail(summary)
```
# Write csv 
```{r}
write.csv(summary, "tables_supplementary_diversity/data/tableS3")
```

